How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-04-15 1.01-beforeyoustart.Rmd
# Changing sections to chapters

### What is a PhD?

@@

### Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

@@

### Criteria used to judge PhD studies
In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university:

@@

### Who are the examiners?
The next question to ask is who the examiners are? Again, different institutions will have different guidelines about who can be an examiner for a PhD. It is likely that most will agree that they should be holders of PhDs, in tenured positions (sometimes universities insist on full professorial appointments for examiners), and with experience of advising PhD students as well as having a proven record of their own publications. Hence, most PhD examiners will likely be senior academic figures. Their views are likely to have been informed by their own experience in academia, their own institutions and systems and prior experience of examining PhDs at other institutions.

@@

### What will you do after your PhD?
It sounds silly to think about what comes after your PhD even as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

#### Doing a PhD actually reduces your chance of employment and lowers your pay
The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification. It will take you so far down an academic rabbit hole, that for an employer there’s very little to use. Most employers want people who understand the nuts and bolts of doing research, but who can apply this knowledge outside of the academic system. Employers who want people with PhDs (and there are a few) generally want professional researchers who are essentially academics. Good work if you can get it, but it tends not to pay well. You might end up only being fit to be an academic, and that life is getting increasingly harder without any end to the expectations of the employers… like the equivalent of doing a PhD every year. Something only very few people will enjoy.

@@

### Think about your career before you start your PhD
The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do. Although this does sound ludicrous, there are many people in academia who did PhDs for this reason, or simply because they were offered an opportunity to do a PhD and felt that they couldn’t turn it down.

@@

### Don't rely on becoming an academic
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs [@afonso2014academia].

### Doing a PhD is really hard work
It’s not a MSc with an extra year. It really is much harder and tougher on you mentally and physically. The chances that you’ll end up jaded and malcontent with the entire academic system are very high (maybe even obligatory). You are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun).

### Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network. It is worth tracking your mental health during your PhD to check that you are not getting into difficulties.

### Any PhD has to be yours
It’s entirely useless to do a PhD for someone else. You must take complete ownership in order to do it. In fact, you need to be obsessed with the PhD subject and really want to do nothing else at all. A PhD is an obsession where you can be unwavering and manic in your fascination with a subject. Very few people really understand what a great privilege this is. If you don’t feel like a total maniac for learning, a PhD is not for you.

### Does it matter where you do your PhD?
Not all universities in the world have a PhD program. This means that you cannot do a PhD in some institutions. Larger institutions are more likely to hold PhD programs, have more potential advisors, and may even have funds for bursaries to conduct PhD studies. It is likely that you are going to need to do some work at the university in order to justify any bursary that they may want to give you. Typically this kind of work is the teaching of undergraduates which is a generally good and positive experience in the world of academia.

@@

### What to look for in an advisor
Throughout this book I will call the person who is the academic in charge of your PhD studies your ‘advisor’. I will not call them your supervisor. The reason for this is because I think it is much better for you to think of this person as giving you advice and not the person who tells you what to do. This comes back to the idea that the PhD is yours. It is not your advisor’s PhD. They will not write it. They will not do all the work that's required for it. That will be yours alone.

@@

### What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisers who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided.

@@

### What makes a good project?
After having made lots of statements about how a project should be your own project the following advice might seem a bit strange. Clearly the work that you're going to be doing will need to fit into the group of the adviser that you've chosen.

@@

### Funding your PhD studies
Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways:

#### The funding that you need to live 
This money is typically called a bursary or stipend depending on where you are in the world. It seems to me that many countries offer bursaries that are not actually enough to live on (e.g. buy food, rent a house etc.). If you're offered a PhD bursary, especially if it's in another country, then do check on the living cost at that location and whether or not the bursary that you are being offered will be sufficient money in order for you to survive. Also, be very careful to check how long the bursary is for. For example, at my university a PhD bursary is given only for 3 years. After this three-year funding period, you will be expected to have submitted your PhD for examination. If you need further funding to live while your PhD is being examined, and before you graduate then you will need to get a job. There is no more money!

@@

#### The cost associated with your studies. 
The money associated with the project that you were doing also needs to come from somewhere. Depending on where in the world you can do these studies they will be different expectations of where this money comes from. This is a very important topic for you and your potential advisor to discuss. During the proposal phase of your PhD you'll be expected to put together a budget to determine the total costs of your studies. In reality you or your advisor should have a ballpark figure of how much this Is going to cost and have some idea about where the money is going to come from.

>>>>

## What is a PhD?

@@

## Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

@@

## Criteria used to judge PhD studies
In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university:

@@

## Who are the examiners?
The next question to ask is who the examiners are? Again, different institutions will have different guidelines about who can be an examiner for a PhD. It is likely that most will agree that they should be holders of PhDs, in tenured positions (sometimes universities insist on full professorial appointments for examiners), and with experience of advising PhD students as well as having a proven record of their own publications. Hence, most PhD examiners will likely be senior academic figures. Their views are likely to have been informed by their own experience in academia, their own institutions and systems and prior experience of examining PhDs at other institutions.

@@

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD even as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

### Doing a PhD actually reduces your chance of employment and lowers your pay
The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification. It will take you so far down an academic rabbit hole, that for an employer there’s very little to use. Most employers want people who understand the nuts and bolts of doing research, but who can apply this knowledge outside of the academic system. Employers who want people with PhDs (and there are a few) generally want professional researchers who are essentially academics. Good work if you can get it, but it tends not to pay well. You might end up only being fit to be an academic, and that life is getting increasingly harder without any end to the expectations of the employers… like the equivalent of doing a PhD every year. Something only very few people will enjoy.

@@

## Think about your career before you start your PhD
The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do. Although this does sound ludicrous, there are many people in academia who did PhDs for this reason, or simply because they were offered an opportunity to do a PhD and felt that they couldn’t turn it down.

@@

## Don't rely on becoming an academic
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs [@afonso2014academia].

## Doing a PhD is really hard work
It’s not a MSc with an extra year. It really is much harder and tougher on you mentally and physically. The chances that you’ll end up jaded and malcontent with the entire academic system are very high (maybe even obligatory). You are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun).

## Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network. It is worth tracking your mental health during your PhD to check that you are not getting into difficulties.

## Any PhD has to be yours
It’s entirely useless to do a PhD for someone else. You must take complete ownership in order to do it. In fact, you need to be obsessed with the PhD subject and really want to do nothing else at all. A PhD is an obsession where you can be unwavering and manic in your fascination with a subject. Very few people really understand what a great privilege this is. If you don’t feel like a total maniac for learning, a PhD is not for you.

## Does it matter where you do your PhD?
Not all universities in the world have a PhD program. This means that you cannot do a PhD in some institutions. Larger institutions are more likely to hold PhD programs, have more potential advisors, and may even have funds for bursaries to conduct PhD studies. It is likely that you are going to need to do some work at the university in order to justify any bursary that they may want to give you. Typically this kind of work is the teaching of undergraduates which is a generally good and positive experience in the world of academia.

@@

## What to look for in an advisor
Throughout this book I will call the person who is the academic in charge of your PhD studies your ‘advisor’. I will not call them your supervisor. The reason for this is because I think it is much better for you to think of this person as giving you advice and not the person who tells you what to do. This comes back to the idea that the PhD is yours. It is not your advisor’s PhD. They will not write it. They will not do all the work that's required for it. That will be yours alone.

@@

## What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisers who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided.

@@

## What makes a good project?
After having made lots of statements about how a project should be your own project the following advice might seem a bit strange. Clearly the work that you're going to be doing will need to fit into the group of the adviser that you've chosen.

@@

## Funding your PhD studies
Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways:

### The funding that you need to live 
This money is typically called a bursary or stipend depending on where you are in the world. It seems to me that many countries offer bursaries that are not actually enough to live on (e.g. buy food, rent a house etc.). If you're offered a PhD bursary, especially if it's in another country, then do check on the living cost at that location and whether or not the bursary that you are being offered will be sufficient money in order for you to survive. Also, be very careful to check how long the bursary is for. For example, at my university a PhD bursary is given only for 3 years. After this three-year funding period, you will be expected to have submitted your PhD for examination. If you need further funding to live while your PhD is being examined, and before you graduate then you will need to get a job. There is no more money!

@@

### The cost associated with your studies. 
The money associated with the project that you were doing also needs to come from somewhere. Depending on where in the world you can do these studies they will be different expectations of where this money comes from. This is a very important topic for you and your potential advisor to discuss. During the proposal phase of your PhD you'll be expected to put together a budget to determine the total costs of your studies. In reality you or your advisor should have a ballpark figure of how much this Is going to cost and have some idea about where the money is going to come from.
